---
output:
  md_document:
    variant: markdown_github
---
<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r, echo = FALSE}

# to render this, use:
if (FALSE) rmarkdown::render("README.Rmd")

suppressPackageStartupMessages({
  library(icd9)
  library(magrittr)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
patientData <- data.frame(
  visitId = c(1000, 1000, 1000, 1000, 1001, 1001, 1002),
  icd9 = c("40201", "2258", "7208", "25001", "34400", "4011", "4011"),
  poa = c("Y", NA, "N", "Y", "X", "Y", "E"),
  stringsAsFactors = FALSE
  )
```
# icd9

[![Build Status](https://travis-ci.org/jackwasey/icd9.svg?branch=master)](https://travis-ci.org/jackwasey/icd9) [![Coverage Status](https://coveralls.io/repos/jackwasey/icd9/badge.svg?branch=master)](https://coveralls.io/r/jackwasey/icd9?branch=master)

# ICD-9 comorbidities, manipulation and validation

## Main Features

 * assignment of patients to high level comorbidities based on admission or discharge ICD-9 codes
     * several mappings of ICD-9 codes to comorbidities are included (Quan, Deyo, Elixhauser, AHRQ)
     * very fast assignment of ICD-9 codes to comorbidities (using C++ internally, with automatic parallel execution using OpenMP when possible)
 * Charlson and Van Walvaren score calculations
 * validation of ICD-9 codes from different annual revisions of ICD-9-CM
 * summarizing ICD-9 codes into groups, and to human-readable descriptions
 * correct conversion between different representations of ICD-9 codes, with and without a decimal point
 * comprehensive test suite to increase confidence in accurate processing of ICD-9 codes.

## New since last CRAN release:
 * further performance increases: 1 million ICD-9 codes assigned to comorbidities in a couple of seconds
 * logical matrix or data.frame for comorbidity output and manipulation
 * see NEWS.md and github [changelog](https://github.com/jackwasey/icd9/commits/master) for more details

## Introduction

Calculate comorbidities, Charlson scores, perform fast and accurate validation, conversion, manipulation, filtering and comparison of ICD-9-CM (clinical modification) codes. ICD-9 codes appear numeric but leading and trailing zeroes, and both decimal and non-decimal "short" format codes exist. The package enables a work flow from raw lists of ICD-9 codes from hospital billing databases to comorbidities. ICD-9 to comorbidity mappings from Quan (Deyo and Elixhauser versions), Elixhauser and AHRQ included. Any other mapping of codes, such as ICD-10, to comorbidities can be used.

## Relevance

ICD-9 codes are still in wide use around the world, particularly in the USA where the ICD-9-CM (Clinical Modification) is in widespread use. ICD-10 and the corresponding ICD-10-CM are imminent, however a vast amount of patient data is recorded with ICD-9 codes of some kind: this package enables their use in R. A common requirement for medical research involving patients is determining new or existing comorbidities. This is often reported in *Table 1* of research papers to demonstrate the similarity or differences of groups of patients. This package is focussed on fast and accurate generation of this comorbidity information from raw lists of ICD-9 codes.


## ICD-9 code types

ICD-9 codes are not numbers, and great care is needed when matching individual codes and ranges of codes. It is easy to make mistakes, hence the need for this package. ICD-9 codes can be presented in *short* 5 character format, or *decimal* format, with a decimal place separating the code into two groups. There are also codes beginning with V and E which have different validation rules. Zeroes after a decimal place are meaningful, so numeric ICD-9 codes cannot be used in most cases. In addition, most clinical databases contain invalid codes, and even decimal and non-decimal format codes in different places. This package primarily deals with ICD-9-CM (Clinical Modification) codes, but should be applicable or easily extendible to the original WHO ICD-9 system.

## Examples

See the vignette and code help for many more. Here's a taste: 

```{r example}
patientData

# reformat input data as needed:
patientData %>% icd9LongToWide # everything works well with magrittr

# get comorbidities:
icd9ComorbidQuanDeyo(patientData)

# find diagnoses present on admission:
icd9FilterPoa(patientData)
```
## Install

The latest version is available in [github](https://github.com/jackwasey/icd9) and can be installed with:

    # install.packages("devtools") # if needed
    devtools::install_github("jackwasey/icd9")

    install.packages("magrittr") # recommended, but not required

The _master_ branch at github should always build and pass all tests and R CMD check, and will be similar or identical to the most recent CRAN release. The CRAN releases are stable milestones. Contributions and bug reports are encouraged.
