---
title: icd Readme
output:
  md_document:
    variant: markdown_github
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: Jack O. Wasey 
---
<!--
Copyright (C) 2014 - 2018  Jack O. Wasey

This file is part of icd.

icd is free software: you can redistribute it and/or modify
it under the terms of the GNU General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

icd is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
GNU General Public License for more details.

You should have received a copy of the GNU General Public License
along with icd. If not, see <http:#www.gnu.org/licenses/>.
-->

<!-- README.md is generated from README.Rmd. Please edit that file and render with 
rmarkdown::render("README.Rmd")
-->
```{r, echo = FALSE}

suppressPackageStartupMessages({
  library(icd)
  library(magrittr)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
patient_data <- data.frame(
  visit_id = c(1000, 1000, 1000, 1000, 1001, 1001, 1002),
  icd9 = as.icd9cm(c("40201", "2258", "7208", "25001", "34400", "4011", "4011")),
  poa = c("Y", NA, "N", "Y", "X", "Y", "E"),
  stringsAsFactors = FALSE
  )
```
# icd
[![CRAN version](https://www.r-pkg.org/badges/version/icd "CRAN Version")](https://cran.r-project.org/package=icd)
[![Travis](https://travis-ci.org/jackwasey/icd.svg?branch=master "Travis Build Status")](https://travis-ci.org/jackwasey/icd)
[![Appveyor](https://ci.appveyor.com/api/projects/status/github/jackwasey/icd "AppVeyor Build Status")](https://ci.appveyor.com/project/jackwasey/icd)
[![codecov.io](https://codecov.io/github/jackwasey/icd/coverage.svg?branch=master)](https://codecov.io/github/jackwasey/icd?branch=master)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/icd)](https://cran.r-project.org/package=icd)
<!-- [![Coverity](https://scan.coverity.com/projects/15671/badge.svg "Coverity Scan Build Status")](https://scan.coverity.com/projects/jackwasey-icd) -->

# ICD-9 and ICD-10 comorbidities, manipulation and validation

## Features

 * find comorbidities of patients based on admission or discharge ICD-9 or ICD-10 codes, e.g. Cancer, Heart Disease
     + several standard mappings of ICD codes to comorbidities are included (Quan, Deyo, Elixhauser, AHRQ)
     + very fast assignment of ICD codes to comorbidities (using matrix multiplication with C and C++ internally)
 * Charlson and Van Walraven score calculations
 * Hierarchical Condition Codes (HCC) from CMS
 * Clinical Classifcations Software (CCS) comorbidities from AHRQ
 * validation of ICD codes from different annual revisions of ICD-9-CM and ICD-10-CM
 * summarizing ICD codes into groups, and to human-readable descriptions
 * correct conversion between different representations of ICD codes, with and without a decimal points, leading and trailing characters (this is not trivial for ICD-9-CM). ICD-9 to ICD-10 cross-walk is not yet implemented
 * comprehensive test suite to increase confidence in accurate processing of ICD codes
 * all internal ICD and comorbidity data is extracted directly from publically available data or code, increasing confidence in the results

## Install

```{r simpleinstall, eval = FALSE, echo = TRUE}
install.packages("icd")
```

## Introduction

```{r pkgdesc, results='asis', echo = FALSE}
cat(packageDescription("icd")$Description)
```

## Relevance

ICD-9 codes are still in heavy use around the world, particularly in the USA where the ICD-9-CM (Clinical Modification) was in widespread use until the end of 2015. ICD-10 has been used worldwide for reporting cause of death for more than a decade. ICD-10-CM is now the primary coding scheme for US hospital admission and discharge diagnoses used for regulatory purposes and billing. A vast amount of patient data is recorded with ICD-9 codes of some kind: this package enables their use in R alongside ICD-10.

## Comorbidities

A common requirement for medical research involving patients is determining new or existing comorbidities. This is often reported in *Table 1* of research papers to demonstrate the similarity or differences of groups of patients. This package is focussed on fast and accurate generation of this comorbidity information from raw lists of ICD-9 codes.

## ICD-9 codes

ICD-9 codes are not numbers, and great care is needed when matching individual codes and ranges of codes. It is easy to make mistakes, hence the need for this package. ICD-9 codes can be presented in *short* 5 character format, or *decimal* format, with a decimal place separating the code into two groups. There are also codes beginning with V and E which have different validation rules. Zeroes after a decimal place are meaningful, so numeric ICD-9 codes cannot be used in most cases. In addition, most clinical databases contain invalid codes, and even decimal and non-decimal format codes in different places. This package primarily deals with ICD-9-CM (Clinical Modification) codes, but should be applicable or easily extendible to the original WHO ICD-9 system.

## ICD-10 codes

ICD-10 has a somewhat simpler format, with consistent use of a letter, then two alphanumeric characters. However, especially for ICD-10-CM, there are a multitude of qualifiers, e.g. specifying recurrence, laterality, which vastly increase the number of possible codes. This package recognizes validity of codes by syntax alone, or whether the codes appear in a canonical list. The current ICD-10-CM master list is the 2016 set. There is no capability of converting between ICD-9 and ICD-10, but comorbidities can be generated from older ICD-9 codes and newer ICD-10 codes in parallel, and the comorbidities can then be compared.

## Examples

See also the vignettes and examples embedded in the help for each function for more. Here's a taste: 

```{r example}
patient_data

# get comorbidities using Quan's application of Deyo's Charlson comorbidity groups
comorbid_charlson(patient_data)

# or go straight to the Charlson scores:
charlson(patient_data)

# get comorbidities based on present-on-arrival diagnoses, use magrittr to flow the data
patient_data %>% filter_poa %>% comorbid_elix
```

Look at the help files for details and examples of almost every function in this package.
```{r how to get further examples, echo = TRUE, eval = FALSE}
?comorbid
?is_valid
```

Note that reformatting from wide to long and back is not as straightforward as using the various Hadley Wickham tools for doing this: knowing the more detailed structure of the data let's us do this better for the case of dealing with ICD codes.

## Advanced

### Source Data and SAS format files

In the spirit of reproducible research, all the R data files in this package can be recreated from source. The size of the source files makes it cumbersome to include them in the R package available on CRAN. Using the github source, you can pull the original data and SAS format files, and rebuild the data; or use the tools provided by this package to update the data using new source data files, e.g. when ICD-10-CM 2017 is released.

### Development version

The latest version is available in [github icd](https://github.com/jackwasey/icd), and can be installed with:
```{r show install, eval = FALSE, echo = TRUE}
    install.packages("devtools")
    devtools::install_github("jackwasey/icd")
```

The _master_ branch at github should always build and pass all tests and R CMD check, and will be similar or identical to the most recent CRAN release. The CRAN releases are stable milestones. Contributions and bug reports are encouraged and essential for this package to remain current and useful to the many people who have installed it.


### Contributing and Building

A substantial amount of code has now been contributed to the package. Contributions of any kind to `icd` are very welcome. See the [GitHub issues page]](https://github.com/jackwasey/icd/issues) to see jobs and feature requests. Documentation, vignettes and examples are very welcome, especially if accompanied by some real-world data.

To build `icd`, `Rcpp` must be compiled from source. This happens automatically on Linux, but on Mac and Windows, the following is required:
`install.packages("Rcpp", type="source")`
to avoid build errors.
