---
title: "icd: Efficient Computation of Comorbidities from ICD Codes Using Sparse Matrix Multiplication in R"
subtitle: "Benchmark replication instructions"
author: "Jack O. Wasey"
date: "`r Sys.Date()`"
output: html_document
---
```{r include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```
### Introduction

The following instructions allow the user to install the dependencies for replicating the benchmarks reported in the accompanying document entitled, "icd: Efficient Computation of Comorbidities from ICD Codes Using Sparse Matrix Multiplication in R"

The benchmarking code is necessarily time-consuming and resource intensive to run, and may take hours to complete. This has been done successfully by the lead developer on hardware ranging from a 2014 laptop, to a 1TB RAM, 72 core server. The benchmarking can be run with or without the biggest and most resource-intensive tests to demonstrate that it functions correctly.

The specifications of the machine used for the reported benchmark results are in the main article.

All the code in this document should be run with the R or shell working directory set to the directory created after decompressing the replication materials archive.

### Quick start

Running `make` in the replication materials directory will complete all the following steps and run abbreviated benchmarks to demonstrate replicability for the benchmarks with a smaller number of iterations. See below to run the complete, very time-consuming benchmarks reported in the article.

### Dependencies

The following script will make sure a CRAN repository is set, create a working library directory named `icd-bench-lib`, and install all the benchmark dependencies into it. It will also attempt to clean up on exit.

```{r}
source("install-dependencies.R")
```

```{r}
```
