---
title: "icd: Efficient Computation of Comorbidities from ICD Codes Using Sparse Matrix Multiplication in R"
subtitle: "Benchmark replication instructions"
author: "Jack O. Wasey"
date: "`r Sys.Date()`"
output: html_document
---
```{r include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction

The following instructions allow the user to install the dependencies for replicating the benchmarks reported in the accompanying document entitled, "icd: Efficient Computation of Comorbidities from ICD Codes Using Sparse Matrix Multiplication in R."

The benchmarking code is necessarily time-consuming and resource intensive to run, and may take hours to complete. This has been done successfully by the lead developer on hardware ranging from a 2014 laptop, to a 1TB RAM, 72 core server. The benchmarking can be run with or without the biggest and most resource-intensive tests to demonstrate that it functions correctly. The specifications of the machine used for the reported benchmark results are in the main article.

All the code in this document should be run with the R or shell working directory set to the directory created after decompressing the replication materials archive. Packages will be installed in the user's library, according to the content of `.libPaths()`, which may be modified by setting the environment variable `R_LIBS_USER` to a temporary library directory before starting R. This may then be safely deleted after completing benchmarking. Packages are installed from CRAN, using the user's current `options("repos")` configuration, if valid, and, failing that, the [R-project cloud repository URL](https://cloud.r-project.org/).

It may be necessary to update installed packages using `update.packages()`. In particular, Rcpp has specific requirements about sometimes needing to recompiled packages that depend on it. See Rcpp documentation for details in event of problems.

# Requirements for replication materials

There are requirements for various R packages beyond those stated in the package `DESCRIPTION`, which are just for running or building the `icd` package. Additional packages are required for replicating the benchmarks.

## Windows

`Rscript` and `make` should be on the path.

## Running the benchmarks

In a shell, set the working directory to the replication materials directory. In the source package structure, this is in `benchmarks/icd-JSS3447-replication`. For replicating the materials submitted to JSS, the compressed archive of the replication materials should be unpacked to a directory, and that chosen directory should be set as the working directory.

Running `make` in the replication materials directory will install dependencies, and run the abbreviated benchmarking code with simulated data of up to 10^3 rows. This small benchmark takes a few seconds. For replication of the full benchmarking presented in the article, `make result7` should be invoked. Be aware, this may take many hours, even on a large multi-core server, since most of the time, the competing packages are bound to a single threads. Benchmarks of intermediate size may be invoked with `make result4`, `make result5`, and make `result6`.

A `dput` dump is made for each benchmarking run, which gives R code which contains the benchmarking results. These results may be used in place of the pre-calculated benchmark results that are in the submitted replication materials for the submitted article.

## Alternative

The preferred method for running the benchmarks is using `make`. Alternatively, the abreviated benchmarks can be run from within R as follows:

```{r}
source("install-dependencies.R")
source("bench-versus.R")
```

This will install dependencies and run the abbreviated benchmarks. To run longer benchmarks, the user may then call the `bench_versus` function. E.g.:

```{r}
bench_versus(4)
```
