---
title: "Calculating comorbidities from medical codes"
author: "Jack O. Wasey"
date: "4/9/2018"
output: 
  pdf_document: 
    keep_tex: yes
vignette: >
  %\VignetteIndexEntry{Calculating comorbidities from medical codes}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---
<!--
Copyright (C) 2014 - 2018  Jack O. Wasey

This file is part of icd.

icd is free software: you can redistribute it and/or modify
it under the terms of the GNU General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

icd is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
GNU General Public License for more details.

You should have received a copy of the GNU General Public License
along with icd. If not, see <http:#www.gnu.org/licenses/>.
-->
```{r setup, echo = FALSE, cache = FALSE}
suppressWarnings({
  suppressPackageStartupMessages({
    loadNamespace("knitr") # for opts_chunk only
    library(icd)
    library(magrittr)
    library(utils)
    })
  })

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

patients_icd9 <- data.frame(
  visit_id = c(1000, 1000, 1000, 1000, 1001, 1001, 1002),
  icd9 = as.icd9(c("40201", "2258", "7208", "25001", "34400", "4011", "4011")),
  poa = c("Y", NA, "N", "Y", "X", "Y", "E"),
  stringsAsFactors = FALSE
  )
```







### AHRQ comorbidity classification

The AHRQ keeps an updated version of the Elixhauser classification of ICD-9-CM and ICD-10-CM codes into comorbidities. They provide the data in the form of SAS code. The names of the comorbidities derived from ICD-9 and ICD-10 codes are the same. Maps contain the ICD code to comorbidity mappings; the functions that apply those mappings are called things like `icd10_comorbid_ahrq`.
```{r ahrq}
#icd9_map_ahrq <- icd:::sas_parse_ahrq() # user doesn't need to do this
names(icd9_map_ahrq)
icd9_map_ahrq$CHF[1:5]
icd10_map_ahrq$CHF[1:5]
```

### Quan
Quan's paper looked at indices using both ICD-10 and ICD-9-CM. Quan generated updated ICD-9-CM codes for all 30 of Elixhauser and all 17 of Charlson/Deyo's comorbidities. Thus there are two 'Quan' comorbidity mappings.
```{r quan elix}
names(icd10_map_quan_deyo)
names(icd10_map_quan_elix)
```

















