---
title: "ICD-10 comorbidities"
author: "Jack O. Wasey"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: html_document
vignette: >
  %\VignetteIndexEntry{ICD-10 comorbidities}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
<!--
Copyright (C) 2014 - 2015  Jack O. Wasey

This file is part of icd9.

icd9 is free software: you can redistribute it and/or modify
it under the terms of the GNU General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

icd9 is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
GNU General Public License for more details.

You should have received a copy of the GNU General Public License
along with icd9. If not, see <http:#www.gnu.org/licenses/>.
-->



```{r setup, include=FALSE}

suppressWarnings({
  suppressPackageStartupMessages({
    loadNamespace("knitr") # for opts_chunk only
    library(icd9)
    library(magrittr)
    library(xtable)
    library(utils)
    })
  })

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

```

Using the fast code assignment in this package isn't dependent on ICD-9 or ICD-9-CM codes. Any codes can be used, and this package now includes some sample ICD-10-CM codes, and mappings from ICD-10 codes to comorbidities as defined by Quan et al. The sample data are from http://www.ustur.wsu.edu/database/ where de-identified diagnoses are given for a few hundred pathology cases relating to uranium exposure.

The sample data is in the 'long' format, i.e. multiple rows per case

```{r uranium-long}
uranium_pathology[1:10, ]
```

We can convert the date from long to wide format easily. Tthis is done quickly in C++, and gives reliable output with factors). It also has other features which make it work well with other functions in this package. However, you may be happier using dplyr and friends for this manipulation.

```{r uranium-wide}
uranium_pathology %>% icd_long_to_wide %>% head
```

Now map these diagnoses to disease groups as defined by Quan et al:
```{r comorbidities}
quan_comorbidities <- icd_comorbid(uranium_pathology, icd10_map_quan_elix)
# see the first few rows and columns:
quan_comorbidities[1:6, 3:10]
```

The ICD-10-CM mappings are recorded a bit differently from the ICD-9-CM mappings in this package. The ICD-9 mappings included all possible permutations of child codes. Since ICD-10 codes contain letters, and are seven characters long, this became impractical. Therefore, the current mappings include only codes for the most recent update of ICD-10-CM. This means that slightly miscoded diseases, e.g. with a spurious additional number, will not be counted.

For example, here are the codes for hypertension with complications from Quan et al:
```{r htncx}
icd10_map_quan_elix$HTNcx
```
