---
title: "Working with World Health Organization ICD codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with World Health Organization ICD codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Surprisingly, there are potential copyright restrictions on redistributing the WHO ICD-9 and ICD-10 definitions, which form the basis of ICD code systems around the world. These definitions are themselves also used as the internationally standard way to codify morbidity and mortality for public health.

`icd` now includes the ability to work with these codes more easily via the companion `icd.data` package. The data is not available in the package so `icd.data` asks the user whether they wish to download the data from the WHO web site, whereupon it is made available just like any other R data, but the name `icd.data::icd10who2016`, which is the latest ICD-10 release from the WHO at the time of writing.

```{r quicklook, eval=FALSE}
icd.data::icd10who2016[1:5, 1:5]
library(icd.data)
summary(icd10who2016)
```
